A highly-available move operation for replicated trees

### 2020-05-04 paper/move-op.tex
Change to use lipics style

\documentclass[acmsmall,anonymous,review]{acmart}
% Recommended options from acmart-pacmpl-template.tex
\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage{tikz}
\usetikzlibrary{arrows.meta}

@@

\setcopyright{none}
\acmJournal{PACMPL}
\acmVolume{}
\acmNumber{OOPSLA}
\acmArticle{}
\acmYear{2020}
\acmMonth{1}
\acmDOI{}
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}

@@

\begin{document}
\title{A highly-available move operation for replicated trees}

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Dominic P.\ Mulligan}
\email{Dominic.Mulligan@arm.com}
\orcid{0000-0003-4643-3541}
\affiliation{%
  \institution{Arm Research}
  \streetaddress{110 Fulbourn Road}
  \city{Cambridge}
  \state{}
  \postcode{CB1 9NJ}
  \country{United Kingdom}
}

\author{Victor B.\ F.\ Gomes}
\email{vb358@cl.cam.ac.uk}
\orcid{0000-0002-2954-4648}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Alastair R.\ Beresford}
\email{arb33@cl.cam.ac.uk}
\orcid{0000-0003-0818-6535}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\begin{abstract}
    Replicated tree data structures are data models supporting distributed filesystems, such as Google Drive and Dropbox, and XML or JSON databases.
    These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
    However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
    Our algorithm requires no synchronous coordination between replicas, making it highly available in the face of network partitions.
    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant, and evaluate the performance of our formally verified implementation in a geo-replicated setting.
\end{abstract}

\maketitle

@@

\begin{figure*}
\centering
\begin{tikzpicture}
  \tikzstyle{time}=[thick,->,gray]
  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
  \node [anchor=east] at (-1.2,3) {Replica 1:};
  \node [anchor=east] at (-1.2,0) {Replica 2:};
  \node [rectangle,draw] (start1) at (0,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start2) at (0,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change1) at (4.5,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change2) at (4.5,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw,inner sep=3mm] (merge1) at (8,3) {?};
  \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
  \draw [time] (change1.east) -- (merge1.west);
  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
  \draw [time] (change2.east) -- (merge2.west);
  \draw [network] (6.0,0) to [out=90,in=270] (6.7,3);
  \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
  \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
  %%%%%
  \node [rectangle,draw] at (-1.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
      \node at (0,1) {$\mathsf{root}$};
      \node (a1) {$A$} child {node (b1) {$B$}} child {node {$C$}};
      \draw (b1.south west) .. controls (-2,-2) and (0,1.5) .. (a1.north);
      \end{tikzpicture}
  };
  \node [rectangle,draw] at (1.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.94,0.03) {(\emph{b})};
      \tikzstyle{level 1}=[sibling distance=12mm]
      \tikzstyle{level 2}=[sibling distance=6mm]
      \node {$\mathsf{root}$}
          child {node {$A$} child {node {$B$}} child {node {$C$}}}
          child {node {$B'$} child {node {$A'$} child {node {$C'$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start) at (4.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (right) at (7.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.93,0.03) {(\emph{d})};
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
\end{tikzpicture}
\caption{Initially, nodes $A$ and $B$ are siblings. Replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$. Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
\Description{Visualisation of trees with concurrent move operations}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
  \label{fig:gdrive-error}
  \Description{Screenshot showing the errors ``Can't remove files with changes pending. Try again after sync is complete.'' and ``An unknown error occurred.''}
\end{figure*}

@@

\Description{The move operation algorithm, implemented in the Isabelle/HOL language (continued in Figure~\ref{fig:code2}).}

@@

\Description{The move operation algorithm (continuation of Figure~\ref{fig:code1}).}

@@

\begin{figure}
  \includegraphics[width=10cm,keepaspectratio=true]{plot.pdf}
  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates. Error bars indicate the minimum and 95th percentile execution times.}
  \label{fig:plots}
  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds. Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
\end{figure}

\begin{figure}
  \includegraphics[width=10cm,keepaspectratio=true]{leader-plot.pdf}
  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region. Note the log scale on both axes.}
  \label{fig:leader-plot}
  \Description{Local CRDT operations take about 50 microseconds, and reach saturation at 600 ops/sec. Leader-based operations take about 150--180ms, and reach saturation at 14,000 ops/sec.}
\end{figure}

@@

\begin{acks}
The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
Our evaluation was conducted using AWS credits from the AWS Educate program.
Thank you to Marc Shapiro for feedback on a draft of this paper.
\end{acks}

\bibliography{references}{}
\end{document}

>>>>

\documentclass[a4paper,anonymous,USenglish]{lipics-v2019}

\usepackage{booktabs} % \toprule etc.
\usepackage{tikz}
\usetikzlibrary{arrows.meta}

@@

@@

\title{A highly-available move operation for replicated trees}

\author{Martin Kleppmann}{University of Cambridge}{mk428@cl.cam.ac.uk}{https://orcid.org/0000-0001-7252-6958}{Supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.}

\author{Dominic P. Mulligan}{Arm Research, Cambridge}{Dominic.Mulligan@arm.com}{https://orcid.org/0000-0003-4643-3541}{}

\author{Victor B. F. Gomes}{Google\footnote{Work done while at the University of Cambridge}}{vb358@cl.cam.ac.uk}{https://orcid.org/0000-0002-2954-4648}{}

\author{Alastair R. Beresford}{University of Cambridge}{arb33@cl.cam.ac.uk}{https://orcid.org/0000-0003-0818-6535}{}

\authorrunning{M. Kleppmann, D.\,P. Mulligan, V.\,B.\,F. Gomes, and A.\,R. Beresford}
\Copyright{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}

\funding{The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528). Our evaluation was conducted using AWS credits from the AWS Educate program.}
\acknowledgements{Thank you to Marc Shapiro for feedback on a draft of this paper.}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003809.10010172</concept_id>
       <concept_desc>Theory of computation~Distributed algorithms</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
   <concept>
       <concept_id>10011007.10010940.10010992.10010998.10010999</concept_id>
       <concept_desc>Software and its engineering~Software verification</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Distributed algorithms}
\ccsdesc[500]{Software and its engineering~Software verification}

\keywords{Conflict-free Replicated Data Types (CRDTs), distributed filesystems, formal verification}

\begin{document}
\maketitle
\begin{abstract}
    Replicated tree data structures are data models supporting distributed filesystems, such as Google Drive and Dropbox, and XML or JSON databases.
    These systems need to support a \emph{move} operation that allows a subtree to be moved to a new location within the tree.
    However, such a move operation is difficult to implement correctly if different replicas can concurrently perform arbitrary move operations, and we demonstrate bugs in Google Drive and Dropbox that arise with concurrent moves.
    In this paper we present an algorithm that handles arbitrary concurrent modifications on trees, while ensuring that the tree structure remains valid (in particular, no cycles are introduced), and guaranteeing that all replicas converge towards the same consistent state.
    Our algorithm requires no synchronous coordination between replicas, making it highly available in the face of network partitions.
    We formally prove the correctness of our algorithm using the Isabelle/HOL proof assistant, and evaluate the performance of our formally verified implementation in a geo-replicated setting.
\end{abstract}

@@

\begin{figure*}
\centering
\begin{tikzpicture}
  \tikzstyle{time}=[thick,->,gray]
  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
  \node [anchor=east] at (-1.2,3) {Replica 1:};
  \node [anchor=east] at (-1.2,0) {Replica 2:};
  \node [rectangle,draw] (start1) at (0,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start2) at (0,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change1) at (4.5,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change2) at (4.5,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw,inner sep=3mm] (merge1) at (8,3) {?};
  \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
  \draw [time] (change1.east) -- (merge1.west);
  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
  \draw [time] (change2.east) -- (merge2.west);
  \draw [network] (6.0,0) to [out=90,in=270] (6.7,3);
  \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
  \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
  %%%%%
  \node [rectangle,draw] at (-1.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
      \node at (0,1) {$\mathsf{root}$};
      \node (a1) {$A$} child {node (b1) {$B$}} child {node {$C$}};
      \draw (b1.south west) .. controls (-2,-2) and (0,1.5) .. (a1.north);
      \end{tikzpicture}
  };
  \node [rectangle,draw] at (1.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.94,0.03) {(\emph{b})};
      \tikzstyle{level 1}=[sibling distance=12mm]
      \tikzstyle{level 2}=[sibling distance=6mm]
      \node {$\mathsf{root}$}
          child {node {$A$} child {node {$B$}} child {node {$C$}}}
          child {node {$B'$} child {node {$A'$} child {node {$C'$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start) at (4.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (right) at (7.5,-3.5) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.93,0.03) {(\emph{d})};
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
\end{tikzpicture}
\caption{Initially, nodes $A$ and $B$ are siblings. Replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$. Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
  \label{fig:gdrive-error}
\end{figure*}

@@

@@

@@

\begin{figure}
  \includegraphics[width=10cm,keepaspectratio=true]{plot.pdf}
  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates. Error bars indicate the minimum and 95th percentile execution times.}
  \label{fig:plots}
\end{figure}

\begin{figure}
  \includegraphics[width=10cm,keepaspectratio=true]{leader-plot.pdf}
  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region. Note the log scale on both axes.}
  \label{fig:leader-plot}
\end{figure}

@@

\bibliographystyle{plainurl}
\bibliography{references}{}
\end{document}
